---
title: "Data Outliers"
author: "Dr. Adrian Correndo"
date: "2026-02-14"
categories: [Agronomy, Statistics, Decision Making]
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-tools: true
editor: source
---

## Why this matters 🧭

During exploratory data analysis (EDA), you will often see observations that look unusual (very high/low yield, odd nutrient values, a point far from the trend). A common impulse is to delete them before modeling.

In agronomy and most sciences, **automatic deletion at EDA is not a good default**. Unusual values can be caused by:

- **Error** (data entry, unit mismatch, sensor failure, mislabeling) 🧾
- **Reality** (true variability: soil/drainage gradients, pest/disease, stand loss, weather, management differences) 🌱🌦️
- **Model mismatch** (a simple model misses structure) 🧠

---

## Core principle ✅

> **“Outlier” ≠ “bad data.”**  
> EDA is the stage to **flag and investigate**, not to “optimize” the dataset.

---

## A practical workflow 🔎➡️📈

| Step | Goal | What to do | Typical output |
|---|---|---|---|
| 1. Screen | Catch obvious issues | Range checks, missingness, duplicates, unit checks | A list of flagged rows |
| 2. Investigate | Decide what’s wrong vs real | Compare against field notes/metadata; check plausibility; cross-check related variables | Classification + notes |
| 3. Model | Fit and diagnose | Fit model; inspect residual patterns; check influence | Diagnostics and influence flags |
| 4. Sensitivity | Test robustness | Refit with documented exclusions / influential points removed; compare estimates | “Do conclusions change?” |
| 5. Report | Be transparent | Document rules; keep raw data; report primary + sensitivity results | Reproducible decisions |

---

## How to classify unusual observations 🧩

::: {.callout-tip title="Three categories to keep straight"}

1. **Confirmed errors** 🧾  
   - wrong units (lbs/ac recorded as kg/ha), decimal shift, impossible values, duplicates, corrupted rows  <br/> 
2. **Protocol deviations** 🚜  
   - plot not harvested, mislabeled sample, equipment failure, known flooding/animal damage   <br/>
3. **Rare but plausible values** 🌾  
   - extreme but possible outcomes due to real field variability   <br/>

:::

---

## When is deletion justified? 🗂️

::: {.callout-note title="Deletion is justified when you can defend it with evidence"}

Reasonable reasons include: <br/>
- **Impossible values** (e.g., negative yield)  <br/>
- **Unit/entry mistakes** confirmed by checks <br/>
- **Duplicate records** <br/>
- **Confirmed protocol violations** documented in metadata/field notes <br/>

**Best practice** <br/>
- Keep a **raw dataset unchanged** <br/>
- Create a **cleaned dataset** with explicit rules <br/>
- Record what changed in a **data-cleaning log** <br/>

:::

---

## When should you NOT delete? ⚠️

::: {.callout-warning title="Avoid deleting points just because they look extreme"}
Do **not** remove observations because:

- they are “far from the mean” <br/>
- they reduce significance or worsen model fit <br/>
- they are hard to explain <br/>
- they might represent true field variability (common in agronomy) <br/>

If removing one or two points flips your conclusion, the key message is usually:
> the inference is fragile (design/model/measurement may need improvement), not that the data should be “fixed.”
:::

---

## Influence vs outliers (model-based thinking) 🎯

An observation can be unusual without changing conclusions. What often matters more is whether it is **influential**.

::: {.callout-note title="Key ideas"}
- **Large residual**: unusual *Y* given *X*
- **High leverage**: unusual *X* (predictor pattern)
- **Influence**: strongly changes fitted coefficients/predictions
:::

---

## Studentized residuals (outliers in Y given X) 📏

A **residual** is $e_i = y_i - \hat{y}_i$. To compare residuals across observations, we standardize them.

**Studentized residuals** account for differences in residual variability and leverage. In practice, the most common diagnostic in R is the *externally studentized* residual:

- `rstudent(fit)` returns the studentized residual for each observation (computed leaving that observation out of the residual SD estimate).

::: {.callout-note title="Interpretation"}
Large absolute studentized residuals indicate observations whose **Y value is unusually far from the model expectation**, given their X values.
:::

::: {.callout-warning title="Rules of thumb (for flagging, not automatic deletion)"}

- $|r_i| > 2$: potentially unusual  

- $|r_i| > 3$: strongly unusual (worth investigating)  

With large datasets, some points will exceed these cutoffs by chance, so treat them as **screens**.
:::

---

## Cook’s distance (influential observations) 🍳

**Cook’s distance** measures how much the fitted model would change if observation \(i\) were removed. It combines:

- how unusual the point is in **Y** (residual size) and

- how unusual the point is in **X** (leverage)

In R:
- `cooks.distance(fit)` computes Cook’s distance for each observation.

::: {.callout-note title="Interpretation"}
A large Cook’s distance means:
> “This observation can have a strong impact on the fitted coefficients and the conclusions.”
:::

::: {.callout-warning title="Rules of thumb (for flagging)"}
Common cutoffs used for screening include:

- $D_i > \frac{4}{n}$ (simple and widely used)  

- $D_i > 1$ (often considered very influential, especially in smaller datasets)

These are not universal thresholds. They are **starting points for investigation**.
:::

---

## Recommended reporting practice 🧾✅

::: {.callout-tip title="A simple, defensible template"}
1. **Primary analysis**: model fitted on the full dataset (after correcting *confirmed errors*).  
2. **Sensitivity analysis**: refit after excluding  
   - confirmed errors/protocol violations (documented), and optionally  
   - highly influential points (diagnostic only)  
3. **Compare**: key coefficients, standard errors, treatment differences, and conclusions.
:::

---

## Example in R 💻

### pacakges

```{r}
library(pacman)
p_load(dplyr, tidyr) # data wrangling
p_load(ggplot2) # figures
p_load(broom, performance)
```

### data
```{r}
# Example:
cornn_data <- read.csv("../week_06/data/cornn_data.csv") %>% 
  mutate(n_trt = factor(n_trt, levels = c("N0","N60","N90","N120","N180","N240","N300")),
         block = factor(block),
         site = factor(site),
         year = factor(year))
```
### model fit
```{r}
# Fit model
fit <- lm(yield_kgha ~ n_trt, data = cornn_data)
```
### Residuals table
```{r}
###
# Create table with residuals and flag suspicious observations
diag_tbl <- augment(fit, data = cornn_data) %>%
  # .std.resid = standardized residuals (broom)
  mutate(
    stud_resid = rstudent(fit),           # studentized residuals (externally studentized)
    cooks_d    = .cooksd,                 # Cook's distance from broom
    flag_stud  = abs(stud_resid) > 3,
    flag_cook  = cooks_d > 4 / n(),
    flagged    = flag_stud | flag_cook
  ) %>%
  arrange(desc(flagged), desc(cooks_d), desc(abs(stud_resid)))

# View only flagged observations (tidy table)
diag_tbl %>%
  filter(flagged) %>%
  select(everything())

```
### Plots 
#### i. studentizided residuals
```{r}
# plot outliers
diag_tbl %>%
  ggplot(aes(x = .fitted, y = stud_resid)) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed") +
  geom_hline(yintercept = 0)+
  geom_point(aes(shape = flagged, fill = flagged), alpha = 0.7) +
  scale_shape_manual(values = c(21,24))+
  scale_fill_brewer(type = "qual", palette = 2, direction = 1)+
  geom_text(data = diag_tbl %>% filter(flagged),
      aes(label = block), # use block as label
          vjust = -0.7,
              size = 3) +
  labs(title = "Potential outliers",
      x = "Fitted values",
      y = "Studentized residuals",
      shape = "flagged"
  ) +
  theme_classic()
```

#### ii. Cook's distance
```{r}
diag_tbl %>%
  mutate(cook_cut = 4 / n()) %>%
  ggplot(aes(x = seq_along(cooks_d), y = cooks_d)) +
  geom_hline(aes(yintercept = cook_cut), linetype = "dashed") +
  geom_point(aes(shape = flagged, fill = flagged), alpha = 0.7) +
  scale_shape_manual(values = c(21,24))+
  scale_fill_brewer(type = "qual", palette = 2, direction = 1)+
  geom_text(data = diag_tbl %>% filter(flagged),
      aes(label = block), # use block as label
          vjust = -0.7,
              size = 3) +
  labs(title = "Potential influence",
      subtitle = "Cook's distance",
      x = "Observation index",
      y = "Cook's distance",
      shape = "flagged"
  ) +
  theme_classic()
```

#### iii. Hybrid option
```{r}
diag_tbl %>%
  ggplot(aes(x = .hat, y = stud_resid)) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed") +
  geom_hline(aes(yintercept = 0), linetype = "solid") +
  geom_point(aes(size = cooks_d, shape = flagged, fill = flagged), alpha = 0.7) +
  scale_shape_manual(values = c(21,24))+
  scale_fill_brewer(type = "qual", palette = 2, direction = 1)+
  geom_text(data = diag_tbl %>% filter(flagged),
      aes(label = block), # use block as label
          vjust = -0.7,
              size = 3) +
  labs(
    title = "Influence map",
    x = "Leverage (hat values)",
    y = "Studentized residuals",
    size = "Cook's D",
    shape = "Flagged"
  ) +
  theme_classic()
```

#### other options
The `performance` package offers a variety of functions to check assumptions and for outliers/influential observations
```{r}
performance::check_model(model = fit, check = c("outliers"))
```
The `car` package also has useful functions for outlier detection and influence diagnostics such as 'outlierTest()' and 'influencePlot()':
```{r}
car::outlierTest(fit)
car::influencePlot(fit)
```


## Treatment options 🛠️
What to do instead of deleting

::: {.callout-tip title="Options in increasing order of intervention"}
1. **Do nothing** (keep the point) if it is plausible and conclusions are robust  
2. **Fix errors** (units, typos) with documentation  
3. **Add structure** (block/site/year random effects, covariates, interactions) if the point reveals missing model structure  
4. **Model non-constant variance** or transform the response if residual spread grows with mean  
5. **Robust methods** if the goal is stable estimation under heavy tails (useful, but interpret carefully)  
:::

---

## Reliable R packages 📦

::: {.callout-note title="Good, reliable tools for diagnostics"}
**Base R (reliable and transparent)**  
- `rstudent()`, `rstandard()`  
- `cooks.distance()`, `hatvalues()`, `dfbetas()`  
- `influence.measures()`  

**performance** (great for teaching)  
- `performance::check_model()` for a consistent diagnostic overview (and it works for many model types)  

**car** (classic and widely used)  
- `car::outlierTest()` (Bonferroni-adjusted test for large residuals in `lm`)  
- `car::influencePlot()` (visual summary of leverage/residual/influence)  

**broom** (tidy output)  
- `broom::augment()` to attach fitted values and residuals back to a dataset  
:::

---

## Final Suggestion 📌

In this course, we do not delete observations during EDA unless they are **confirmed errors** or **clear protocol violations**. Instead, we flag unusual points, fit models, evaluate influence, and use sensitivity analyses. Any exclusions must be documented and reproducible.  

## Packages used for outliers & influence diagnostics 📦

- **broom**: provides `augment()` to attach observation-level diagnostics (e.g., fitted values, residuals, leverage, Cook’s distance when available) back onto your data frame for tidy workflows. [https://cran.r-project.org/package=broom](https://cran.r-project.org/package=performance)

- **performance**: provides `check_model()` for a compact, teaching-friendly set of diagnostic plots (assumptions + influential observations) that works across many model classes.  [https://cran.r-project.org/package=performance](https://cran.r-project.org/package=performance)

- **car**: classic applied regression tools; includes dedicated functions commonly used for outliers/influence such as `outlierTest()` and `influencePlot()` (bubble plot of studentized residuals vs leverage with Cook’s distance). [https://cran.r-project.org/package=car](https://cran.r-project.org/package=car)

## References 📚

- Cook, R. D. (1977). *Detection of influential observation in linear regression*. **Technometrics**, 19(1), 15–18. [https://doi.org/10.1080/00401706.1977.10489493](https://doi.org/10.1080/00401706.1977.10489493)

- Weisberg, S. (2014). *Applied Linear Regression* (4th ed.). Wiley. **Chapter 9: Regression Diagnostics**. [https://www.stat.purdue.edu/~qfsong/teaching/525/book/Weisberg-Applied-Linear-Regression-Wiley.pdf](https://www.stat.purdue.edu/~qfsong/teaching/525/book/Weisberg-Applied-Linear-Regression-Wiley.pdf)

- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning* (2nd ed.). Springer. **Chapter 3: Linear Regression** (see the section on “Potential Problems”). [https://www.statlearning.com](https://www.statlearning.com)

