---
title: "Reading tabular data in R"
author: "Dr. Adrian Correndo"
date: "2026-01-21"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    toc-location: left
    code-tools: true
categories: files, csv, excel, spreadsheet, readr, readxl, openxlsx
execute:
  echo: true
  eval: true
editor: source
---

# Introduction 📌

In real projects, a big part of “data science" is simply getting data into R reliably and reproducibly. CSV (comma separated values) and Excel are the most common formats you’ll see in agriculture, industry, and research workflows.

This short note covers:  <br/>
- 📄 how to read CSV (readr). <br/>
- 📊 how to read Excel (readxl, openxlsx).  <br/>
- 🗂️ how to read many files in a folder. <br/>
- 🧾 how to read many sheets in an excel file.  <br/>
- 🧹 common cleaning options like skipping rows, header rows, and missing values  <br/>

# Libraries 📦

```{r}
library(pacman)
p_load(dplyr, purrr, stringr, tibble) # data wrangling
p_load(readr, readxl, openxlsx) # to read files
```

## Quick comparison of main packages

📄 CSV / delimited text  
- Package: readr  
- ✅ Strengths: fast, consistent parsing, clear messages, good type control (col_types)  
- ⚠️ Notes: great default for tidyverse-style workflows  

📊 Excel (.xlsx)  
- Package: readxl  
- ✅ Strengths: simple reading of Excel, minimal dependencies  
- ⚠️ Notes: read-only (does not write Excel)  

📊 Excel (.xlsx)  
- Package: openxlsx  
- ✅ Strengths: read + write, more control over excels/sheets  
- ⚠️ Notes: great when you must write .xlsx outputs  

Rule of thumb:
- 📄 CSV: start with readr::read_csv()
- 📊 Excel reading: start with readxl::read_excel()
- 📝 Excel reading + writing or more control: use openxlsx

# Reading CSV files 📄

## Basic read

```{r}
df <- readr::read_csv("data/my_file.csv")
df
```

## Common options
missing values, quieter output

```{r}
df <- readr::read_csv(
  file = "data/my_file.csv",
  na = c("", "NA", "N/A", "na", ".")
)
```

## Control column types 🎯
Recommended for reproducibility.
```{r}
df <- readr::read_csv(file = "data/my_file.csv") |> 
    mutate(loc  = as.character(loc),
           year  = as.integer(year),
           yield = as.numeric(yield))

```

## Skip lines at the top  ⏭️
For example for metadata, notes, extra header rows...

```{r}
df <- readr::read_csv(
  file = "data/my_file.csv",
  skip = 1
)
```

## Skip blank lines 🕳️
Often helpful with messy CSV exports

```{r}
df <- readr::read_csv(
  file = "data/my_file.csv",
  skip_empty_rows = TRUE
)
```

# Reading Excel files (.xlsx) 📊

## Basic read (first sheet by default)

```{r}
df <- readxl::read_excel(path="data/my_excel.xlsx")
head(df)
```

## Read a specific sheet
by name or position

```{r}
df1 <- readxl::read_excel(path="data/my_excel.xlsx", sheet = "Sheet1")
df2 <- readxl::read_excel(path="data/my_excel.xlsx", sheet = 2)
```

## Skip rows ⏭️
For example, title row + empty row

```{r}
df <- readxl::read_excel(
  path = "data/my_excel.xlsx",
  sheet = 1,
  skip  = 2
)
```

## Diagnose 🔎
For example, "empty first row" or "headers" not where expected"

```{r}
df_rows <- readxl::read_excel(
  path = "data/my_excel.xlsx",
  sheet     = 1,
  n_max     = 6, # number of rows of data
  col_names = FALSE # remove headers row
)
df_rows
```

Then set skip based on what you see (commonly skip = 1 or skip = 2) and re-read.

## Reading Excel with openxlsx 🛠
Nice when you also write excel files later

```{r}
df_sr1 <- openxlsx::read.xlsx(
  xlsxFile = "data/my_excel.xlsx",
  sheet    = 1,
  startRow = 1 # speficies where headers are located
)
head(df_sr1)
```

### Skip rows by moving startRow ⏭️:

```{r}
df_sr3 <- openxlsx::read.xlsx(
  xlsxFile = "data/my_excel.xlsx",
  sheet    = 1,
  startRow = 3
)
slice_head(df_sr3)
```

# Reading multiple CSV files 🗂️

This is extremely common when we have one file per site, year, trial, etc.

## Read + stack 🧱
This is call row-bind

```{r}
folder <- "data"

files <- list.files(path = folder, pattern = "\\.csv$", full.names = TRUE)

all_data <- files |>
  setNames(basename(files)) |> # here we name by file basename
  # Here we read and stack using map_dfr
  purrr::map_dfr(
    \(f) readr::read_csv(f), # here we read each file
    .id = "source_file" # this creates a new column with the file name
  )

head(all_data)
```

## Add variables from filename  🏷️
For example year, or site

```{r}
all_data <- files |>
  purrr::map_dfr(
    # here we read each file
    \(f) readr::read_csv(f) |> 
      dplyr::mutate(source_file = basename(f))  ) |> # add filename as column
  dplyr::mutate(year = stringr::str_extract(source_file, "\\d{4}")) 
# here we extract a 4-digit year from the filename

all_data %>% 
  group_by(source_file) %>%
  slice_head()
```

## Keep them as a named list 📚

```{r}
csv_list <- files |>
  setNames(basename(files)) |>
  purrr::map(\(f) readr::read_csv(f))

names(csv_list)
```

# Reading multiple Excel files 🗂️📊

```{r}
folder <- "data"
files  <- list.files(path = folder, pattern = "\\.xlsx$", full.names = TRUE)

all_xlsx <- files |>
  purrr::map_dfr(
    \(f) readxl::read_excel(f, sheet = 1) |> # read first sheet from each
      dplyr::mutate(source_file = basename(f)) # add filename as column
  )

all_xlsx %>% 
  group_by(source_file) %>%
  slice_head()
```

# Reading multiple sheets 🧾🧾🧾

## With readxl

```{r}
path <- "data/my_excel.xlsx"

sheets <- readxl::excel_sheets(path)

all_sheets <- sheets |>
  setNames(sheets) |>
  purrr::map_dfr(
    \(sh) readxl::read_excel(path, sheet = sh) |>
      dplyr::mutate(sheet = sh)
  )

all_sheets  %>% 
  group_by(sheet) %>%
  slice_head()
```

## With openxlsx 🛠️

```{r}
path <- "data/my_excel.xlsx"

sheets <- openxlsx::getSheetNames(path)

all_sheets <- sheets |>
  setNames(sheets) |>
  purrr::map_dfr(
    \(sh) openxlsx::read.xlsx(path, sheet = sh) |>
      dplyr::mutate(sheet = sh)
  )

all_sheets %>% 
  group_by(sheet) %>%
  slice_head()
```

## Footnotes in Excel 📝
If a sheet has footnotes at the bottom, you can often remove them after import by filtering rows:

```{r}
df <- readxl::read_excel(path="data/my_excel.xlsx") |>
  dplyr::filter(!is.na(yield))   # example: keep rows that have yield
```

# Summary

- Use readr::read_csv() for most CSV workflows (fast + consistent).
- Use readxl::read_excel() for quick Excel imports.
- Use openxlsx when you need more Excel control or want to write .xlsx outputs.
- For multiple files or sheets, combine list.files()/excel_sheets() with purrr::map_dfr() to create one clean dataset.